---
title: "TurboRibo"
subtitle: "Filtering, Normalization, Clustering"
author: "Kennedi Todd"
date: "03/19/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Setup
## Set working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

## Load libraries
```{r libraries, message=FALSE, warnings=FALSE}
# load libraries
library(cowplot)     # plot_grid()
library(dplyr)       # left_join()
library(ggplot2)     # ggplot()
library(gridExtra)   # grid.arrange()
library(parallel)    # detectCores()
library(rtracklayer) # import()
library(scCustomize) # Merge_Seurat_List()
library(Seurat)      # Read10X_h5()
library(stringr)     # str_match()
```

## Variables and functions
```{r set_variables_and_thresholds}
# variables
samples <- c("PP_4","PP_11","PP_13","PP_14","PP_15","PP_19","PP_22","PP_23")

# thresholds
nCount.min <- 1000
nCount.max <- 20000
nFeature.min <- 600
complexity.cutoff <- 0.8
mt.cutoff <- 1

# single cell functions
source("../../../../functions/Kennedi_single_cell_functions_v5.R")

# save function
saveToPDF <- function(...) {
    d = dev.copy(pdf,...)
    dev.off(d)
}
```

# Load data
## Merge h5
```{r read_h5, warning=FALSE}
prefix <- "../../counts/"
suffix <- "/outs/filtered_feature_bc_matrix.h5"

if (file.exists("../../rObjects/raw_h5.rds")) {
  mouse <- readRDS("../../rObjects/raw_h5.rds")
} else {
  
  # loop to create objects
  for (i in samples) {
    print(i)
    obj <- CreateSeuratObject(Read10X_h5(paste0(prefix,i,suffix)))
    assign(i, obj)
  }
  
  # merge objects
  mouse <- merge(x = PP_4,
                 y = c(PP_11,
                       PP_13, PP_14,
                       PP_15, PP_19,
                       PP_22, PP_23),
                 add.cell.ids = samples, 
                 project = "TurboRibo snRNAseq")
  
  saveRDS(mouse, "../../rObjects/raw_h5.rds")
}

# preview
mouse

# cleanup and save
remove(PP_4, PP_11, PP_13, PP_14,
       PP_15, PP_19, PP_22, PP_23, obj)
gc()
```

## Annotation
- Annotation file was downloaded from 10x Genomics \
  + refdata-gex-mm10-2020-A \
  + provider: GENCODE \
  + description: evidence-based annotation of the mouse genome (GRCm38), version M23 (Ensembl 98) \
- Gm* genes are originally annotated by MGI and the *Rik genes are annotated by RIKEN \
```{r read_annotation}
# read in annotation file, GENCODE GRCm38 version M23 (Ensembl 98)
if (file.exists("../../rObjects/annotation.rds")) {
  genes <- readRDS("../../rObjects/annotation.rds")
} else {
  gtf.file <- "/research/labs/neurology/fryer/projects/references/mouse/refdata-gex-mm10-2020-A/genes/genes.gtf"
  genes <- rtracklayer::import(gtf.file)
  genes <- as.data.frame(genes)
  genes <- genes[genes$type == "gene",]
  saveRDS(genes, "../../rObjects/annotation.rds")
}
```

## Metadata columns
```{r add_metadata_columns}
# read meta
meta <- read.delim2("../../refs/mika_meta.tsv", sep = "\t")
colnames(meta) <- c("sequence_id","cohort","genotype","sex","i7_index","i5_index")
meta$mouse_id <- as.numeric(str_match(meta$sequence_id, "PP_([0-9]+)")[,2])
meta$sample <- paste(meta$genotype, meta$sex, meta$mouse_id, sep = ".")
meta$sample <- gsub("Female","F",meta$sample)
meta$sample <- gsub("Male","M",meta$sample)

# parse sample
barcodes <- colnames(mouse)
pattern <- "(.+)_[ACGT]+-(\\d+)"
sequence.id <- str_match(barcodes, pattern)[,2]
table(sequence.id)
mouse$sequence_id <- factor(sequence.id)

# join meta tables
mouse.meta <- mouse@meta.data
mouse.meta$barcodes <- rownames(mouse.meta)
mouse.meta <- left_join(mouse.meta, meta, by = "sequence_id")
rownames(mouse.meta) <- mouse.meta$barcodes
mouse@meta.data <- mouse.meta
remove(meta,mouse.meta)

# Set new idents
Idents(mouse) <- "sample"

# cell.complexity
mouse$cell.complexity <- log10(mouse$nFeature_RNA) / log10(mouse$nCount_RNA)

# percent.mt
mt.genes <- genes[genes$seqnames == "chrM",13]
mouse$percent.mt <- PercentageFeatureSet(mouse, features = mt.genes)
mt.genes

# percent.ribo
# ribosomal proteins begin with 'Rps' or 'Rpl' in this annotation file
# mitochondrial ribosomes start with 'Mrps' or 'Mrpl'
gene.names <- genes$gene_name
ribo <- gene.names[grep("^Rp[sl]", gene.names)]
mt.ribo <- gene.names[grep("^Mrp[sl]", gene.names)]
ribo.combined <- c(mt.ribo,ribo)
mouse$percent.ribo.protein <- PercentageFeatureSet(mouse, features = ribo.combined)
ribo.combined

# percent.hb
# percent.hb - hemoglobin proteins begin with 'Hbb' or 'Hba' for mouse
hb.genes <- gene.names[grep("^Hb[ba]-", gene.names)]
mouse$percent.hb <- PercentageFeatureSet(mouse, features = hb.genes)
hb.genes

# nuclear fraction
# init
samples <- c("PP_4","PP_11","PP_13","PP_14","PP_15","PP_19","PP_22","PP_23")
prefix <- "../../refs/"
all.barcodes <- NULL
all.nuc.frac <- NULL
# loop
for (sample in samples) {
  file <- read_tsv(paste0(prefix, sample, "_nuclear_fraction.tsv"), 
                   col_names = FALSE, skip = 1)
  colnames(file) <- c("barcode","nuclear_fraction")
  file$barcode <- paste0(sample, "_", file$barcode)
  all.barcodes <- c(all.barcodes, file$barcode)
  all.nuc.frac <- c(all.nuc.frac, file$nuclear_fraction)
}
# add
all.equal(rownames(mouse@meta.data), all.barcodes)
mouse$nuclear_fraction <- all.nuc.frac * 100
remove(file)
```

# Pre-filtering QC
## Number of cells
```{r prefiltering_cells_per_sample}
# Visualize the number of cell counts per sample
data <- as.data.frame(table(mouse$sample))
colnames(data) <- c("sample","frequency")

ncells1 <- ggplot(data, aes(x = sample, y = frequency, fill = sample)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  #scale_fill_manual(values = sample_colors) + 
  scale_y_continuous(breaks = seq(0,15000, by = 5000), limits = c(0,15000)) +
  ggtitle("Raw: cells per sample") +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))
ncells1
```

## Density plots
```{r prefiltering_density, warning=FALSE}
# Visualize nCount_RNA
den1 <- ggplot(mouse@meta.data,
       aes(color = sample,
           x = nCount_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #scale_color_manual(values = sample_colors) +
  #scale_fill_manual(values = sample_colors) +
  xlab("nCount_RNA") +
  ylab("Density") +
  theme(legend.position =  "none") +
  geom_vline(xintercept = nCount.min) +
  geom_vline(xintercept = nCount.max) +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize nFeature_RNA
den2 <- ggplot(mouse@meta.data,
       aes(color = sample,
           x = nFeature_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("nFeature_RNA") +
  ylab("Density") +
  geom_vline(xintercept = nFeature.min) +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize cell complexity
# Quality cells are usually above 0.85
den3 <- ggplot(mouse@meta.data,
       aes(color = sample,
           x = cell.complexity,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("Cell Complexity (log10(nFeature/nCount))") +
  ylab("Density") +
  geom_vline(xintercept = complexity.cutoff) +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize percent.mt
den4 <- ggplot(mouse@meta.data,
       aes(color = sample,
           x = percent.mt,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(n.breaks = 4) +
  geom_vline(xintercept = mt.cutoff) +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("% Mitochondrial Genes") +
  ylab("Density") +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize percent.ribo.protein
den5 <- ggplot(mouse@meta.data,
       aes(color = sample,
           x = percent.ribo.protein,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #geom_vline(xintercept = ribo.cutoff) +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("% Ribosomal Protein Genes") +
  ylab("Density") +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize percent.hb
den6 <- ggplot(mouse@meta.data,
       aes(color = sample,
           x = percent.hb,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #geom_vline(xintercept = hb.cutoff) +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("% Hemoglobin Genes") +
  ylab("Density") +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Arrange graphs in grid
plots <- list(den1,den2,den3,den4,den5,den6)
layout <- rbind(c(1,4),c(2,5),c(3,6))
grid <- grid.arrange(grobs = plots, layout_matrix = layout)
```

```{r save_prefiltering_density,echo=FALSE,eval=FALSE}
# save
grid <- grid.arrange(grobs = plots, layout_matrix = layout)
path <- paste0("../../results/filtering_QC/density_curves_raw")
saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
dev.off()

# cleanup
remove(den1,den2,den3,den4,den5,den6,layout,grid,plots,data)
```

## Violin plots
```{r prefiltering_violins, warning=FALSE}
# nFeature, nCount, and cell.complexity violins
v1 <- VlnPlot(mouse,
              features = c("nFeature_RNA", "nCount_RNA","cell.complexity"),
              ncol = 3,
              group.by = 'sample',
              #cols = sample_colors,
              pt.size = 0)
v1

#  percent violins
v2 <- VlnPlot(mouse,
              features = c("percent.mt","percent.ribo.protein","percent.hb"),
              ncol = 3,
              group.by = 'sample',
              #cols = sample_colors,
              pt.size = 0)
v2
```

```{r save_prefiltering_violins, echo=FALSE, eval=FALSE}
# save v1
v1
path <- paste0("../../results/filtering_QC/violins_nFeature_nCount_complexity_raw")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 4)
dev.off()

# save v2
v2
path <- paste0("../../results/filtering_QC/violins_percent_raw")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 4)
dev.off()

# cleanup
remove(v1,v2)
```

## Scatter plots
```{r prefiltering_scatter, warning=FALSE}
s1 <- ggplot(
  mouse@meta.data,
  aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
	scale_x_log10() +   	
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = nCount.min) + 
  geom_hline(yintercept = nFeature.min) + 
  facet_wrap(~sample) +
  scale_colour_gradient(low = "gray90", high = "black", limits =c(0,100))
s1

s2 <- FeatureScatter(mouse,
                     feature1 = "nCount_RNA",
                     feature2 = "percent.mt",
                     group.by = 'sample',
                     #cols = sample_colors,
                     shuffle = TRUE)
s2
```

```{r save_prefiltering_scatter, echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE}
# save
s1
path <- paste0("../../results/filtering_QC/scatter_nFeature_vs_nCount_raw")
saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
dev.off()

s2
path <- paste0("../../results/filtering_QC/scatter_percentMT_vs_nCount_raw")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 6)
dev.off()

# cleanup
remove(s1,s2)
```

# Filtering
## Cell-level filtering
```{r cell_filtering}
# filter
mouse.filtered <- subset(mouse,
                        subset = (nCount_RNA > nCount.min) &
                          (nCount_RNA < nCount.max) &
                          (nFeature_RNA > nFeature.min) &
                          (cell.complexity > complexity.cutoff) &
                          (percent.mt < mt.cutoff))

# print cells removed
print(paste0(dim(mouse)[2] - dim(mouse.filtered)[2]," cells removed"))
```

## Gene-level filtering
Remove lowly expressed genes.  We will keep genes that have at least 1 count in 10 cells.
```{r gene_filtering, warning=FALSE, message=FALSE}
# filter genes
mouse.filtered <- JoinLayers(mouse.filtered)
counts <- GetAssayData(object = mouse.filtered, layer = "counts")
nonzero <- counts > 0  # produces logical
keep <- Matrix::rowSums(nonzero) >= 10  # sum the true/false
counts.filtered <- counts[keep,]  # keep certain genes

# overwrite mouse.filtered
mouse.filtered <- CreateSeuratObject(counts.filtered, 
                                     meta.data = mouse.filtered@meta.data)

# print features removed
print(paste0(dim(counts)[1] - dim(counts.filtered)[1], " features removed"))
```


```{r remove_specific_genes}
# remove mt.genes
counts <- GetAssayData(object = mouse.filtered, layer = "counts")
keep <- !rownames(counts) %in% mt.genes # false when mt.gene
counts.filtered <- counts[keep,]

# overwrite mouse.filtered
mouse.filtered <- CreateSeuratObject(counts.filtered,
                                     meta.data = mouse.filtered@meta.data)

# print features removed
print(paste0(dim(counts)[1] - dim(counts.filtered)[1], " features removed"))

# cleanup data
remove(mouse,counts,counts.filtered,nonzero)
```

```{r save_filtered_object, echo=FALSE, eval=FALSE}
saveRDS(mouse.filtered, "../../rObjects/filtered.rds")
#mouse.filtered <- readRDS("../../rObjects/filtered.rds")
```

# Post-filtering QC
## Number of cells
```{r postfiltering_cells_per_sample}
# Visualize the number of cell counts per sample
data <- as.data.frame(table(mouse.filtered$sample))
colnames(data) <- c("sample","frequency")

ncells2 <- ggplot(data, aes(x = sample, y = frequency, fill = sample)) + 
  geom_col() +
  theme_classic() +
  geom_text(aes(label = frequency), 
            position=position_dodge(width=0.9), 
            vjust=-0.25) +
  #scale_fill_manual(values = sample_colors) + 
  scale_y_continuous(breaks = seq(0,15000, by = 5000), limits = c(0,15000)) +
  ggtitle("Filtered: cells per sample") +
  theme(legend.position =  "none") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1))

# Arrange graphs in grid
plots <- list(ncells1,ncells2)
layout <- rbind(c(1),c(2))
grid <- grid.arrange(grobs = plots, layout_matrix = layout)
```

```{r save_cells_per_sample,echo=FALSE,eval=FALSE}
# save
grid <- grid.arrange(grobs = plots, layout_matrix = layout)
path <- "../../results/filtering_QC/cells_per_sample_before_and_after_filtering"
saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
dev.off()

# cleanup
remove(ncells1,ncells2,plots,layout,grid,data)
```

## Density plots
```{r postfiltering_density, warning=FALSE}
# Visualize nCount_RNA
den1 <- ggplot(mouse.filtered@meta.data,
       aes(color = sample,
           x = nCount_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #scale_color_manual(values = sample_colors) +
  #scale_fill_manual(values = sample_colors) +
  xlab("nCount_RNA") +
  ylab("Density") +
  theme(legend.position =  "none") +
  geom_vline(xintercept = nCount.min) +
  geom_vline(xintercept = nCount.max) +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize nFeature_RNA
den2 <- ggplot(mouse.filtered@meta.data,
       aes(color = sample,
           x = nFeature_RNA,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("nFeature_RNA") +
  ylab("Density") +
  geom_vline(xintercept = nFeature.min) +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize cell complexity
# Quality cells are usually above 0.85
den3 <- ggplot(mouse.filtered@meta.data,
       aes(color = sample,
           x = cell.complexity,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("Cell Complexity (log10(nFeature/nCount))") +
  ylab("Density") +
  geom_vline(xintercept = complexity.cutoff) +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize percent.mt
den4 <- ggplot(mouse.filtered@meta.data,
       aes(color = sample,
           x = percent.mt,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_continuous(n.breaks = 4) +
  geom_vline(xintercept = mt.cutoff) +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("% Mitochondrial Genes") +
  ylab("Density") +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize percent.ribo.protein
den5 <- ggplot(mouse.filtered@meta.data,
       aes(color = sample,
           x = percent.ribo.protein,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #geom_vline(xintercept = ribo.cutoff) +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("% Ribosomal Protein Genes") +
  ylab("Density") +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Visualize percent.hb
den6 <- ggplot(mouse.filtered@meta.data,
       aes(color = sample,
           x = percent.hb,
           fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  #geom_vline(xintercept = hb.cutoff) +
  #scale_color_manual(values = sample_colors) +
  theme(legend.position =  "none") +
  #scale_fill_manual(values = sample_colors) +
  xlab("% Hemoglobin Genes") +
  ylab("Density") +
  theme(legend.key.size = unit(0.25, 'cm'), legend.title = element_text(size=9))

# Arrange graphs in grid
plots <- list(den1,den2,den3,den4,den5,den6)
layout <- rbind(c(1,4),c(2,5),c(3,6))
grid <- grid.arrange(grobs = plots, layout_matrix = layout)
```

```{r save_postfiltering_density, echo=FALSE, eval=FALSE, warning=FALSE}
# save
grid <- grid.arrange(grobs = plots, layout_matrix = layout)
path <- "../../results/filtering_QC/density_curves_filtered"
saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
dev.off()

# cleanup
remove(den1,den2,den3,den4,den5,den6,plots,layout,grid)
```

## Violin plots
```{r postfiltering_violins}
# nFeature, nCount, and cell.complexity violins
v3 <- VlnPlot(mouse.filtered,
              features = c("nFeature_RNA", "nCount_RNA","cell.complexity"),
              ncol = 3,
              group.by = 'sample',
              #cols = sample_colors,
              pt.size = 0)
v3

#  percent violins
v4 <- VlnPlot(mouse.filtered,
              features = c("percent.mt","percent.ribo.protein","percent.hb"),
              ncol = 3,
              group.by = 'sample',
              #cols = sample_colors,
              pt.size = 0)
v4
```

```{r save_postfiltering_violins, echo=FALSE, eval=FALSE}
# save
v3
path <- paste0("../../results/filtering_QC/violins_nFeature_nCount_complexity_filtered")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 4)
dev.off()

# save
v4
path <- paste0("../../results/filtering_QC/violins_percent_filtered")
saveToPDF(paste0(path, ".pdf"), width = 12, height = 4)
dev.off()

# cleanup
remove(v3,v4)
```

## Scatter plots
```{r postfiltering_scatter}
s3 <- ggplot(
  mouse.filtered@meta.data,
  aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
	scale_x_log10() +   	
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = nCount.min) + 
  geom_hline(yintercept = nFeature.min) + 
  facet_wrap(~sample) +
  scale_colour_gradient(low = "gray90", high = "black", limits =c(0,100))
s3

s4 <- FeatureScatter(mouse.filtered,
                     feature1 = "nCount_RNA",
                     feature2 = "percent.mt",
                     group.by = 'sample',
                     #cols = sample_colors,
                     shuffle = TRUE)
s4
```

```{r save_postfiltering_scatter,echo=FALSE,eval=FALSE}
# save
s3
path <- paste0("../../results/filtering_QC/scatter_nFeature_vs_nCount_filtered")
saveToPDF(paste0(path, ".pdf"), width = 8, height = 8)
dev.off()

s4
path <- "../../results/filtering_QC/scatter_nCount_vs_percentMT_filtered"
saveToPDF(paste0(path, ".pdf"), width = 6, height = 6)
dev.off()

# cleanup
remove(s3,s4)
```

## Box plot
```{r boxplot}
# Visualize the distribution of genes detected per cell via boxplot
b1 <- ggplot(mouse.filtered@meta.data,
       aes(x = sample, 
           y = log10(nFeature_RNA), 
           fill=sample)) + 
  geom_boxplot() + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  theme(plot.title = element_text(hjust = 0.5, face="bold")) +
  ggtitle("Unique Genes / Cell / Sample") +
  #scale_color_manual(values = sample_colors) +
  #scale_fill_manual(values = sample_colors) +
  xlab("Sample")
b1
```

```{r save_boxplot,echo=FALSE,eval=FALSE}
# save
b1
path <- "../../results/filtering_QC/boxplot_nFeature_per_sample"
saveToPDF(paste0(path, ".pdf"), width = 8, height = 6)
dev.off()

# cleanup
remove(b1)
```

## Top transcripts
- Gm42418: Rn18s-rs5
```{r top_transcripts}
df <- data.frame(gene_name = rownames(mouse.filtered))
df$rsum <- rowSums(x = mouse.filtered, slot = "counts")
df <- df[order(df$rsum, decreasing = TRUE),]
rownames(df) <- 1:nrow(df)
head(df, 10)
```

```{r save_top_transcripts,echo=FALSE,eval=FALSE}
write.table(df, 
            "../../results/filtering_QC/top_transcripts.tsv",
            quote = FALSE,
            row.names = TRUE)
```

# Unwanted variation
## Cell cycle
```{r cell_cycle, eval=FALSE}
markers <- "/research/labs/neurology/fryer/m214960/PLX5622_mice/refs/cell_cycle_markers.tsv"
path <- "../../results/unwanted_variation"
mouse.filtered[["phase"]] <- cell_cycle_QC(obj = mouse.filtered,
                                           markersPath = markers,
                                           species = "mouse",
                                           outDir = path)
```

## Mitochondria
[Evaluating effects of mitochondrial expression](https://hbctraining.github.io/scRNA-seq_online/lessons/06_SC_SCT_normalization.html)
```{r mitochondria_factor, eval=FALSE}
mouse.filtered[["mito.factor"]] <- mitochondria_QC(obj = mouse.filtered,
                                                   outDir = path)
```

# Cluster
## SCTransform
- SCTransform method is a more accurate method of normalizing, estimating the 
variance of the raw filtered data, and identifying the most variable genes. \
- Variation in sequencing depth (total nCount_RNA per cell) is normalized using 
a regularized negative binomial model. \
- If there are other sources of uninteresting variation it can be included. \
- SCTransform output can generate large R objects/variables in terms of memory. \
- Since we have a large dataset, we will adjust the limit for allowable object 
sizes within R (Default is 500 * 1024 ^ 2 = 500 Mb). \
```{r split_object}
# split
mouse.filtered[["RNA"]] <- split(mouse.filtered[["RNA"]], 
                                 f = mouse.filtered$sample)
# transform
mouse.filtered <- SCTransform(mouse.filtered, verbose = FALSE)
```


- NOTE: By default, after normalizing, adjusting the variance, and regressing out 
uninteresting sources of variation, SCTransform will rank the genes by residual 
variance and output the 3000 most variant genes. If the dataset has larger cell 
numbers, then it may be beneficial to adjust this parameter higher using the 
variable.features.n argument. Additionally the last line of output specifies “Set default assay to SCT”. \
- It is suggested to not regress out batch, and instead use a data integration method like Harmony \

## PCA
```{r run_pca, warning=FALSE, message=FALSE}
# run PCA on the merged object
mouse.filtered <- RunPCA(object = mouse.filtered, assay = "SCT")
```

```{r plot_pca}
# Reset idents and levels
DefaultAssay(mouse.filtered) <- "SCT"
Idents(mouse.filtered) <- "sample"

# Plot PCA
pca1 <- DimPlot(mouse.filtered,
                reduction = "pca",
                split.by = "sample",
                group.by = "sample",
                ncol = 4)
pca1
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save pca1
pca1
path <- paste0("../../results/clustering_QC/PCA_sample")
saveToPDF(paste0(path, ".pdf"), width = 10, height = 10)
dev.off()

# cleanup
remove(pca1)
```

## Integrate
```{r}
# integrate
#mouse.integrated <- IntegrateLayers(object = mouse.filtered,
#                                    method = "HarmonyIntegration")

# re-join layers
#mouse.integrated[["RNA"]] <- JoinLayers(mouse.integrated[["RNA"]])

# plot harmony
#h1 <- DimPlot(mouse.integrated,
#              reduction = "harmony",
#              split.by = "sample",
#              group.by = "sample",
#              ncol = 4)
#h1
```

```{r,echo=FALSE,eval=FALSE,message=FALSE,warning=FALSE}
# save h1
#h1
#path <- paste0("../../results/clustering_QC/harmony_sample")
#saveToPDF(paste0(path, ".pdf"), width = 10, height = 10)
#dev.off()

# cleanup
#remove(h1)
```

## UMAP
To overcome the extensive technical noise in the expression of any single gene 
for scRNA-seq data, Seurat assigns cells to clusters based on their PCA scores 
derived from the expression of the integrated most variable genes, with each PC 
essentially representing a “metagene” that combines information across a 
correlated gene set. Determining how many PCs to include in the clustering step 
is therefore important to ensure that we are capturing the majority of the 
variation, or cell types, present in our dataset.
```{r run_UMAP, message=FALSE, warning=FALSE}
# re-join layers
mouse.filtered[["RNA"]] <- JoinLayers(mouse.filtered[["RNA"]])

# Run UMAP
mouse.filtered <- RunUMAP(mouse.filtered,
                          dims = 1:15,
                          reduction = "pca",
                          n.components = 3) # set to 3 to use with VR

# plot UMAP
DimPlot(mouse.filtered,
        shuffle = TRUE)
```

## Clusters
Seurat uses a graph-based clustering approach, which embeds cells in a graph 
structure, using a K-nearest neighbor (KNN) graph (by default), with edges drawn 
between cells with similar gene expression patterns. Then, it attempts to 
partition this graph into highly interconnected ‘quasi-cliques’ or ‘communities’ 
[Seurat - Guided Clustering Tutorial]. \

We will use the FindClusters() function to perform the graph-based clustering. 
The resolution is an important argument that sets the “granularity” of the 
downstream clustering and will need to be optimized for every individual experiment. 
For datasets of 3,000 - 5,000 cells, the resolution set between 0.4-1.4 generally 
yields good clustering. Increased resolution values lead to a greater number of 
clusters, which is often required for larger datasets. \

The FindClusters() function allows us to enter a series of resolutions and will 
calculate the “granularity” of the clustering. This is very helpful for testing 
which resolution works for moving forward without having to run the function for 
each resolution. \

```{r find_neighbors, message=FALSE, warning=FALSE}
# Determine the K-nearest neighbor graph
mouse.unannotated <- FindNeighbors(object = mouse.filtered,
                                   assay = "SCT",
                                   reduction = "pca",
                                   dims = 1:15)

# Determine the clusters for various resolutions
mouse.unannotated <- FindClusters(object = mouse.unannotated,
                                  algorithm = 1, # 1 = Louvain
                                  resolution = seq(0.1,0.8,by=0.1))
mouse.unannotated$seurat_clusters <- mouse.unannotated$SCT_snn_res.0.1
```

## Explore resolutions
```{r explore_resolutions}
# 0.1
DimPlot(mouse.unannotated,
        group.by = "SCT_snn_res.0.1",
        label = TRUE)

# 0.2
DimPlot(mouse.unannotated,
        group.by = "SCT_snn_res.0.2",
        label = TRUE)

# 0.3
DimPlot(mouse.unannotated,
        group.by = "SCT_snn_res.0.3",
        label = TRUE)

# 0.4
DimPlot(mouse.unannotated,
        group.by = "SCT_snn_res.0.4",
        label = TRUE)

# 0.5
DimPlot(mouse.unannotated,
        group.by = "SCT_snn_res.0.5",
        label = TRUE)
```

```{r save_final_object,echo=FALSE,eval=TRUE}
# Save seurat object
saveRDS(mouse.unannotated, "../../rObjects/pass1_unannotated.rds")
```

