---
title: "TurboRibo"
subtitle: "Cluster Annotation"
author: "Kennedi Todd"
date: "03/19/2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

# Setup
## Working directory
```{r working_directory}
knitr::opts_knit$set(root.dir = ".")
```

## Libraries
```{r libraries, message=FALSE, warning=FALSE}
# load packages
library(Seurat)     
library(SeuratObject)
library(dplyr)        # ungroup()
library(ggtree)       # BuildClusterTree()
library(gridExtra)    # grid.arrange()
library(gtools)       # smartbind()
library(parallel)     # detectCores()
library(plotly)       # plot_ly()  # %>%
```

## Variables and functions
```{r set_variables_and_thresholds}
# single cell functions
source("../../../../functions/Kennedi_single_cell_functions_v5.R")

# save function
saveToPDF <- function(...) {
    d = dev.copy(pdf,...)
    dev.off(d)
}
```

## Read data
```{r read_object}
# read obj
mouse.unannotated <- readRDS("../../rObjects/pass1_unannotated.rds")

# set params
DefaultAssay(mouse.unannotated) <- "RNA"
Idents(mouse.unannotated) <- "seurat_clusters"
mouse.unannotated <- NormalizeData(mouse.unannotated)

# preview
mouse.unannotated
```

# Unannotated QC
## UMAP
```{r unannotated_umap}
mouse.unannotated$seurat_clusters <- mouse.unannotated$SCT_snn_res.0.1
Idents(mouse.unannotated) <- "seurat_clusters"
cluster_colors <- c("black","chocolate4","tan","gray30","gray80","red","orange",
                           "gold","yellow","green","forestgreen","cyan","steelblue",
                           "blue","purple","pink","magenta","salmon")

u1 <- DimPlot(object = mouse.unannotated,
              reduction = "umap",
              shuffle = TRUE,
              repel = TRUE,
              label = TRUE,
              cols = cluster_colors)
u1

u2 <- DimPlot(object = mouse.unannotated,
              reduction = "umap",
              shuffle = TRUE,
              repel = TRUE,
              dims = c(2,3),
              cols = cluster_colors,
              label = TRUE)
u2

sample_colors <- c("gray","red","gold","cyan","blue","green","forestgreen","purple")
u3 <- DimPlot(object = mouse.unannotated,
              reduction = "umap",
              shuffle = TRUE,
              repel = TRUE,
              raster = FALSE,
              group.by = "sample",
              cols = sample_colors)
u3
```

```{r save_unannotated_umap, echo=FALSE}
u1
path <- paste0("../../results/UMAP/unannotated_clusters_colored_by_clusters")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 5)
dev.off()

u3
path <- paste0("../../results/UMAP/unannotated_clusters_colored_by_sample")
saveToPDF(paste0(path, ".pdf"), width = 6, height = 4)
dev.off()
```

## Feature plots
```{r feature_plots_unannotated, message=FALSE, warning=FALSE}
# UMAP percent.mt
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "percent.mt") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))

# UMAP percent.ribo
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "percent.ribo.protein") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))

# UMAP percent.hb
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "percent.hb") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))

# UMAP nCount
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "nCount_RNA") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))

# UMAP nFeature
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "nFeature_RNA") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))

# UMAP cell.complexity
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "cell.complexity") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))

# UMAP Eno2 expression
FeaturePlot(mouse.unannotated,
            reduction = "umap", 
            features = "Ttr") + 
  scale_colour_gradientn(colours = c("blue","lightblue","yellow","orange","red"))
```

## Violins
```{r}
VlnPlot(mouse.unannotated,
        features = "nCount_RNA",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        cols = cluster_colors,
        features = "nFeature_RNA",
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        cols = cluster_colors,
        features = "percent.mt",
        split.by = "seurat_clusters")
```

## Cells per cluster
```{r cells_per_cluster}
# Cells per sample per cluster
sample_ncells <- FetchData(mouse.unannotated, 
                     vars = c("ident", "sample")) %>%
  dplyr::count(ident,sample) %>%
  tidyr::spread(ident, n)
sample_ncells

# Cells per sex per cluster
sex_ncells <- FetchData(mouse.unannotated, 
                     vars = c("ident", "sex")) %>%
  dplyr::count(ident,sex) %>%
  tidyr::spread(ident, n)
sex_ncells
```

## Gene histogram
```{r histogram}
# User params
goi <- "Rbfox1"
threshold <- 1

# Subset data
log2.threshold <- log2(threshold + 0.01)
counts.df <- FetchData(mouse.unannotated, vars = goi)
colnames(counts.df) <- "counts"
log2.counts.df <- log2(counts.df + 0.01)

# Histogram
title <- paste0(goi, "\nnCount_RNA > ", threshold)
hist1 <- ggplot(counts.df, aes(x = counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = title, x=NULL, y=NULL) +
  xlab(paste0(goi, " nCount_RNA")) + ylab("# of Samples") + theme_bw() +
  geom_vline(xintercept = threshold, col = "blue") +
  annotate("rect", 
           xmin = -Inf,
           xmax = threshold, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="chocolate4") +
  annotate("rect", 
           xmin = threshold,
           xmax = Inf, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="deepskyblue")

# Histogram log transformed
hist2 <- ggplot(log2.counts.df, aes(x = counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = title, x=NULL, y=NULL) +
  xlab(paste0("Log2(",goi, " nCount_RNA)")) + ylab("# of Samples") + theme_bw() +
  geom_vline(xintercept = log2.threshold, col = "blue") +
  annotate("rect", 
           xmin = -Inf,
           xmax = log2.threshold, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="chocolate4") +
  annotate("rect", 
           xmin = log2.threshold,
           xmax = Inf, 
           ymin = 0, 
           ymax=Inf, 
           alpha=0.2, 
           fill="deepskyblue")

# plot
plots1 <- list(hist1,hist2)
layout1 <- rbind(c(1),c(2))
grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)
```

## Percent gene
```{r percent_gene}
# user define variable
goi <- "Rbfox1"

# Extract counts data
DefaultAssay(mouse.unannotated) <- "RNA"
Idents(mouse.unannotated) <- "seurat_clusters"
geneData <- FetchData(mouse.unannotated,
                      vars = goi,
                      slot = "counts")
geneData <- geneData > 1
table(geneData)
mouse.unannotated$Expression <- geneData

FetchData(mouse.unannotated,
          vars = c("ident", "Expression")) %>%
  dplyr::count(ident, Expression) %>%
  tidyr::spread(ident, n)

# Plot
mouse.unannotated@meta.data %>%
  group_by(seurat_clusters, Expression) %>%
  dplyr::count() %>%
  group_by(seurat_clusters) %>%
  dplyr::mutate(percent = 100*n/sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x=seurat_clusters,y=percent, fill=Expression)) +
  geom_col() +
  ggtitle(paste0("Percentage of cells with > 1 counts for ", goi)) +
  theme(axis.text.x = element_text(angle = 90))
```

## Cluster tree
- Cluster trees are helpful in deciding what clusters to merge. \
```{r cluster_tree_unannotated, message=FALSE, warning=FALSE}
mouse.unannotated <- BuildClusterTree(object = mouse.unannotated,
                                     dims = 1:15,
                                     reorder = FALSE,
                                     reorder.numeric = FALSE)
tree <- mouse.unannotated@tools$BuildClusterTree
tree$tip.label <- paste0(tree$tip.label)

ggtree::ggtree(tree, aes(x, y)) +
  scale_y_reverse() +
  ggtree::geom_tree() +
  ggtree::theme_tree() +
  ggtree::geom_tiplab(offset = 1) +
  ggtree::geom_tippoint(color = cluster_colors[1:length(tree$tip.label)], shape = 16, size = 5) +
  coord_cartesian(clip = 'off') +
  theme(plot.margin = unit(c(0,2.5,0,0), 'cm'))
```

# Potential Markers
## Astrocytes/Bergmann glia
```{r}
VlnPlot(mouse.unannotated,
        features = "Aqp4",
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Gja1",
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Slc1a3",
        split.by = "seurat_clusters")
```

## Endothelial cells
- Flt1: vascular endothelial growth factor receptor (VEGFR) family \
```{r}
VlnPlot(mouse.unannotated,
        features = "Flt1",
        cols = cluster_colors,
        split.by = "seurat_clusters")
```

## Fibroblasts
```{r fibroblasts, warning=FALSE}
VlnPlot(mouse.unannotated,
        features = "Col1a1",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Col1a2",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Dcn",
        cols = cluster_colors,
        split.by = "seurat_clusters")
```

## Microglia
```{r}
VlnPlot(mouse.unannotated,
        features = "C1qb",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "P2ry12",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Tmem119",
        cols = cluster_colors,
        split.by = "seurat_clusters")
```

## Neurons/NSCs
```{r}
VlnPlot(mouse.unannotated,
        features = "Gad1",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Gad2",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Gabra6",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Syt6",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Stmn2",
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Sox11",
        cols = cluster_colors,
        split.by = "seurat_clusters")
```

## Oligodendrocytes/OPCs
```{r}
VlnPlot(mouse.unannotated,
        features = "Mbp", 
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Mog", 
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Olig1", 
        cols = cluster_colors,
        split.by = "seurat_clusters")
```

## Pericytes & SMCs
```{r}
VlnPlot(mouse.unannotated,
        features = "Acta2", # actin alpha 2, smooth muscle
        cols = cluster_colors,
        split.by = "seurat_clusters")
VlnPlot(mouse.unannotated,
        features = "Rgs5", #
        cols = cluster_colors,
        split.by = "seurat_clusters")
```

## Automatically detect markers
```{r find_all_markers, eval=FALSE}
Idents(mouse.unannotated) <- "seurat_clusters"
markers <- SeuratWrappers::RunPrestoAll(
  object = mouse.unannotated,
  assay = "RNA",
  slot = "counts",
  only.pos = FALSE
)

markers <- markers[markers$p_val_adj < 0.01,]

write.table(markers, 
            paste0("../../results/markers/auto_find_markers_adjpval_0.01.tsv"),
            quote = FALSE,
            row.names = FALSE)
```

```{r subset_markers}
# more stringent filtering
all.markers <- markers[markers$p_val_adj < 0.01,]

# compare 
table(all.markers$cluster)

# subset
cluster0 <- all.markers[all.markers$cluster == 0,]
cluster1 <- all.markers[all.markers$cluster == 1,]
cluster2 <- all.markers[all.markers$cluster == 2,]
cluster3 <- all.markers[all.markers$cluster == 3,]
cluster4 <- all.markers[all.markers$cluster == 4,]
cluster5 <- all.markers[all.markers$cluster == 5,]
cluster6 <- all.markers[all.markers$cluster == 6,]
cluster7 <- all.markers[all.markers$cluster == 7,]
cluster8 <- all.markers[all.markers$cluster == 8,]
cluster9 <- all.markers[all.markers$cluster == 9,]
cluster10 <- all.markers[all.markers$cluster == 10,]
cluster11 <- all.markers[all.markers$cluster == 11,]
cluster12 <- all.markers[all.markers$cluster == 12,]
cluster13 <- all.markers[all.markers$cluster == 13,]
cluster14 <- all.markers[all.markers$cluster == 14,]
cluster15 <- all.markers[all.markers$cluster == 15,]
cluster16 <- all.markers[all.markers$cluster == 16,]
cluster17 <- all.markers[all.markers$cluster == 17,]
cluster18 <- all.markers[all.markers$cluster == 18,]
```

# Cluster Annotations
## Cluster 0:
```{r cluster0}
VlnPlot(mouse.unannotated,
        features = cluster0$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 1:
```{r cluster1}
VlnPlot(mouse.unannotated,
        features = cluster1$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 2: 
```{r cluster2}
VlnPlot(mouse.unannotated,
        features = cluster2$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 3: 
```{r cluster3}
VlnPlot(mouse.unannotated,
        features = cluster3$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 4:
```{r cluster4}
VlnPlot(mouse.unannotated,
        features = cluster4$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 5:
```{r cluster5, echo=FALSE, eval=FALSE}
VlnPlot(mouse.unannotated,
        features = cluster5$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 6:
```{r cluster6, message=FALSE}
VlnPlot(mouse.unannotated,
        features = cluster6$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 7: 
```{r cluster7}
VlnPlot(mouse.unannotated,
        features = cluster7$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 8:
```{r cluster8}
VlnPlot(mouse.unannotated,
        features = cluster8$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 9: Fibroblasts
```{r cluster9}
VlnPlot(mouse.unannotated,
        features = cluster9$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 10:
```{r cluster10}
VlnPlot(mouse.unannotated,
        features = cluster10$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 11:
```{r cluster11}
VlnPlot(mouse.unannotated,
        features = cluster11$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 12: Endothelial
```{r cluster12}
VlnPlot(mouse.unannotated,
        features = cluster12$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 13: 
```{r cluster13}
VlnPlot(mouse.unannotated,
        features = cluster13$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Cluster 15: Endothelial?
```{r cluster15}
VlnPlot(mouse.unannotated,
        features = cluster15$gene[1:20],
        cols = cluster_colors,
        stack = TRUE,
        flip = TRUE,
        split.by = "seurat_clusters")
```

## Merge cluster names
```{r assign_identities, eval=FALSE, echo=FALSE}
# rename all identities
mouse.annotated <- RenameIdents(object = mouse.unannotated,
                                "0" = "Macrophages",
                                "1" = "Endothelial",
                                "2" = "Fibroblasts",
                                "3" = "T cells",
                                "4" = "B cells",
                                "5" = "Dendritic cells",
                                "6" = "B cells",
                                "7" = "Choroid",
                                "8" = "Neutrophils",             
                                "9" = "NK cells",          
                                "10" = "Unknown 1",                    
                                "11" = "Unknown 2",           
                                "12" = "Schwann cells",                 
                                "13" = "Unknown 3")                   
mouse.annotated$annotated_clusters <- Idents(mouse.annotated)

mouse.annotated$annotated_clusters <- factor(mouse.annotated$annotated_clusters,
                                             levels = c("B cells",
                                                        "Choroid",
                                                        "Dendritic cells",
                                                        "Endothelial",
                                                        "Fibroblasts",
                                                        "Macrophages",
                                                        "Neutrophils",
                                                        "NK cells",
                                                        "Schwann cells",
                                                        "T cells",
                                                        "Unknown 1",
                                                        "Unknown 2",
                                                        "Unknown 3"))

# set ident
Idents(mouse.annotated) <- "annotated_clusters"
```

## Annotated UMAP
```{r annotated_umap, eval=FALSE, echo=FALSE}
# set colors
cluster_colors <- c("red","orange","gold","green","forestgreen","cyan",
                         "steelblue","blue","purple","pink","chocolate4","gray",
                         "black")

# umap
umap <- DimPlot(object = mouse.annotated, 
        reduction = "umap",
        repel = TRUE,
        group.by = "annotated_clusters",
        cols = cluster_colors)
umap
```

```{r save_annotated_umap, echo=FALSE, eval=FALSE}
umap
path <- paste0("../results/UMAP/annotated_clusters")
saveToPDF(paste0(path, ".pdf"), width = 8, height = 4)
dev.off()
```

```{r save_pass1_annotated_object, echo=FALSE, eval=FALSE}
# save object
saveRDS(mouse.annotated, "../rObjects/pass1_annotated.rds")
```

## Markers
- Itgam = Cd11b
```{r marker_violin, eval=FALSE, echo=FALSE}
goi <- c("Cd19","Fcrl1",     # B cells
         "Enpp2", "Folr1",   # Choroid
         "Itgax", "H2-DMb1", # DCs
         "Flt1", "Pecam1",   # Endothelial
         "Col1a1", "Col1a2", # Fibroblasts
         "C1qa", "C1qb",     # Macrophages
         "Itgam",            # Neutrophils
         "Nkg7", "Klrb1c",   # NK cells
         "Kcna1", "Plp1",    # Schwann cells
         "Cd3e", "Cd3g")     # T cells)    

v <- VlnPlot(mouse.annotated,
             features = goi,
             split.by = "annotated_clusters",
             flip = TRUE,
             stack = TRUE,
             cols = cluster_colors)
v
```

```{r save_marker_violin, eval=FALSE, echo=FALSE}
v
path <- "../results/markers/cluster_markers_violin"
saveToPDF(paste0(path, ".pdf"), width = 15, height = 10)
dev.off()
```

# Shiny App
- scDRcell is the function used to plot the first UMAP
- inpConf is an argument of scDRcell, inpConf = sc1conf
- infConf gets subset by UI column based on inpdrX which is the user's input for x-axis
- inpdrX is used to set the x-axis label
```{r shiny_app, eval=FALSE}
# create new object
shiny.obj <- mouse.annotated
shiny.obj@assays$RNA@var.features <- shiny.obj@assays$SCT@var.features

# set default params
DefaultAssay(shiny.obj) <- "RNA"
Idents(shiny.obj) <- "annotated_clusters"

# create config
names <- colnames(shiny.obj@meta.data)
names <- names[c(27,28,2:26)]
sc.config <- createConfig(obj = shiny.obj,
                          meta.to.include = names)

# change wd
setwd("../../results/all_clusters_pass1/")

# output shiny app folder
makeShinyApp(obj = shiny.obj, 
             scConf = sc.config, 
             gene.mapping = TRUE,
             shiny.title = "All Clusters Lax Thresholds")

# manual config edits
setwd("../../results/all_clusters_pass1/")
sc1conf <- readRDS("shinyApp/sc1conf.rds")
cluster_hex <- "#B5B9BA|#3385BB|#40BBFF|#84BF96|#7F9D55|#F57C7C|#E42622|#FBB268|#FE8D19|#DE9E83|#A6CEE3|#9D7BBA|#977899"
sc1conf[2,4] <- cluster_hex
group_hex <- "#B89B74|#725E47|#B15928|#F7590B|#A6CEE3|#34B1E8|#ED8F47|#F4CC25"
sc1conf[18,4] <- group_hex
saveRDS(sc1conf,"shinyApp/sc1conf.rds")
```


